---
title: "flowchart"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 5
    number_sections: true
package: "`flowchart`"
vignette: >
  %\VignetteIndexEntry{flowchart}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(flowchart)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
```

# Overview 

`flowchart` is a package for drawing participant flow diagrams directly from a dataframe using tidyverse. It provides a set of functions that can be combined with `|>` to create all kinds of flowcharts from a dataframe in an easy way:

- `as_fc()` transforms a dataframe into a `fc` object that can be manipulated by the package.

- `fc_split()` splits a flowchart according to the different values of a column in the dataframe.

- `fc_filter()` creates a filtered box from the flowchart, based on the evaluation of an expression in the dataframe.

- `fc_merge()` combines horizontally two different flowcharts.

- `fc_stack()` combines vertically two different flowcharts.

- `fc_modify()` allows to modify the parameters of the flowchart which are stored in `.$fc`.

- `fc_draw()` draws the flowchart created by the previous functions.

- `fc_export()` allows to export the flowchart drawn to the desired format.

# Installation

```{r eval=FALSE}
install.packages("flowchart")
```

# Example dataset

We will use the built-in datasets in the package:`clinic_patient` and `clinic_visit`. These datasets belong to the same simulated clinical trial, one containing the information per patient and the other per visit. This study simulates a clinical trial in which subjects have been randomised into a control and treatment group after meeting some inclusion criteria: over 18 years of age and signed informed consent. These subjects are then followed for two visits and observations are collected on a biomarker.

```{r}
data(clinic_patient)
data(clinic_visit)

# Per patient dataset
str(clinic_patient)

# Per visit dataset
str(clinic_visit)
```

# Basic operations

The first step is to initialise the flowchart with `as_fc`. The last step, if we want to visualise the created flowchart, is to draw the flowchart with `fc_draw`. In between we can combine the functions `fc_split`., `fc_filter`, `fc_merge`, `fc_stack` with the operator pipe (`|>`) to create complex flowchart structures.

## Initialize

```{r}
fc <- clinic_patient |> 
  as_fc() 

class(fc)
str(fc)
```

The `fc` object created is a list containing the tibble of the dataframe associated with the flowchart and the tibble that stores the flowchart parameters.

Alternatively, if a dataframe is not available, we can initialize a flowchart using the `N =` argument manually specifying the number of rows:

```{r}
fc2 <- as_fc(N = 230)
```

## Draw

```{r}
fc |> 
  fc_draw()
```

## Filter 

We can filter the flowchart using `fc_filter()` specifying the logic in which the filter is to be applied. For example, we want to show the number of patients included in the study, that is, those over the age of 18 who have signed an informed consent. 

```{r fig.width = 6, fig.height = 5}
clinic_patient |> 
  as_fc(label = "Patients included") |> 
  fc_filter(age >= 18 & consent == "Yes", label = "Patients included", show_exc = TRUE) |> 
  fc_draw()
```

See 'Modify function arguments' for more information on the `label=` and `show_exc=` arguments. 

Alternatively, if the column to filter is not available, we can use the `N =` argument to manually specify the number of rows of the resulting filter:

```{r fig.width = 6, fig.height = 5}
clinic_patient |> 
  as_fc(label = "Patients included") |> 
  fc_filter(N = 200, label = "Patients included", show_exc = TRUE) |> 
  fc_draw()
```

## Split

We can split the flowchart into groups using `fc_split()` specifying the grouping variable. The function will split the flowchart into as many categories as the specified variable has:

```{r fig.width = 6, fig.height = 5}
clinic_patient |> 
  filter(!is.na(group)) |> 
  as_fc(label = "Patients included") |> 
  fc_split(group) |> 
  fc_draw()
```

Alternatively, if the column to split is not available, we can use the `N =` argument to manually specify the number of rows in each group of the resulting split:

```{r fig.width = 6, fig.height = 5}
clinic_patient |> 
  filter(!is.na(group)) |> 
  as_fc(label = "Patients included") |> 
  fc_split(N = c(100, 100), label = c("Control", "Treatment")) |> 
  fc_draw()
```

# Combine

`fc_merge()` and `fc_stack()` allow you to combine different flowcharts horizontally or vertically. This is very useful when you need to combine flowcharts generated from different dataframes, as shown here. 

## Merge

We can combine different flowcharts horizontally using `fc_merge()`. For example, we might want to represent the flow of patients included in the study next to the flow of their corresponding visits:

```{r fig.width = 8}
# Create first flowchart for patients
fc1 <- clinic_patient |> 
  filter(!is.na(group)) |> 
  as_fc(label = "Patients included") |> 
  fc_split(group)

# Create second flowchart for visits
fc2 <- clinic_visit |> 
  filter(!is.na(group)) |> 
  as_fc(label = "Number of visits") |> 
  fc_split(group) 

list(fc1, fc2) |> 
  fc_merge() |> 
  fc_draw()
```

## Stack

We can combine different flowcharts vertically using `fc_stack()`. For example, we might want to combine in one flowchart the flow of patients included in the study and the flow of their corresponding visits, which come from different dataframes:

```{r warning = FALSE, fig.width = 6, fig.height = 5}
# Create first flowchart for patients
fc1 <- clinic_patient |> 
  filter(!is.na(group)) |> 
  as_fc(label = "Patients included") |> 
  fc_split(group)

# Create second flowchart for visits
fc2 <- clinic_visit |> 
  filter(!is.na(group)) |> 
  as_fc(hide = TRUE) |> 
  fc_split(group, label = c("Number of visits (Control)", "Number of visits (Treatment)"), text_pattern = "{label}\n {n}") 

list(fc1, fc2) |> 
  fc_stack() |> 
  fc_draw()
```

See 'Modify function arguments' for more information on the `hide=` and `text_pattern=` arguments.

# Customize output

We can customize the flowchart either with the arguments provided by each function in the process of creating it, or directly in the final output using the function `modify_fc`. 

## Modify function arguments

Arguments common to `as_fc()`, `fc_filter()` and `fc_split()`, to customise the appearance of the boxes created at each step:

<table>
 <thead>
  <tr>
   <th style="text-align:left;"> Argument </th>
   <th style="text-align:left;"> Description </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> `label=` </td>
   <td style="text-align:left;"> modify the label. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `text_pattern=` </td>
   <td style="text-align:left;"> modify the pattern of the text. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `text_color=` </td>
   <td style="text-align:left;"> modify the color of the text. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `text_fs=` </td>
   <td style="text-align:left;"> modify the font size of the text. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `bg_fill=` </td>
   <td style="text-align:left;"> modify the background color of the box. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `border_color=` </td>
   <td style="text-align:left;"> modify the border color of the box. </td>
  </tr>
</tbody>
</table>

`as_fc()` arguments:

<table>
 <thead>
  <tr>
   <th style="text-align:left;"> Argument </th>
   <th style="text-align:left;"> Description </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> `hide=` </td>
   <td style="text-align:left;"> hide the first initial box created by this function. </td>
  </tr>
</tbody>
</table>

`fc_filter()` arguments:

<table>
 <thead>
  <tr>
   <th style="text-align:left;"> Argument </th>
   <th style="text-align:left;"> Description </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> `sel_group=` </td>
   <td style="text-align:left;"> apply the filter only in the specified groups (if data is grouped). </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `show_exc=` </td>
   <td style="text-align:left;"> show the box with the excluded rows that do not match the filter. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `direction_exc=` </td>
   <td style="text-align:left;"> change the direction of the exclusion box (left or right). </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `label_exc=` </td>
   <td style="text-align:left;"> modify the label of the exclusion box. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `text_pattern_exc=` </td>
   <td style="text-align:left;"> modify the pattern of the exclusion box. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `text_color_exc=` </td>
   <td style="text-align:left;"> modify the color of the text in the exclusion box. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `text_fs_exc=` </td>
   <td style="text-align:left;"> modify the font size of the text in the exclusion box. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `bg_fill_exc=` </td>
   <td style="text-align:left;"> modify the background color of the exclusion box. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `border_color_exc=` </td>
   <td style="text-align:left;"> modify the border color of the exclusion box. </td>
  </tr>
</tbody>
</table>

`fc_split()` arguments:

<table>
 <thead>
  <tr>
   <th style="text-align:left;"> Argument </th>
   <th style="text-align:left;"> Description </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> `sel_group=` </td>
   <td style="text-align:left;"> split the flowchart only in the specified groups (if data is grouped). </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `na.rm=` </td>
   <td style="text-align:left;"> omit the missing values in the grouping variable. </td>
  </tr>
</tbody>
</table>

`fc_draw()` arguments are heredited from `arrow`: 

<table>
 <thead>
  <tr>
   <th style="text-align:left;"> Argument </th>
   <th style="text-align:left;"> Description </th>
  </tr>
 </thead>
<tbody>
  <tr>
   <td style="text-align:left;"> `arrow_angle=` </td>
   <td style="text-align:left;"> angle of the arrow head in degrees. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `arrow_length=` </td>
   <td style="text-align:left;"> unit specifying the length of the arrow head. </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `arrow_ends=` </td>
   <td style="text-align:left;"> specify the ends of the line to draw the arrow head (last/first/both). </td>
  </tr>
  <tr>
   <td style="text-align:left;"> `arrow_type=` </td>
   <td style="text-align:left;"> whether the arrow head should be a closed triangle. </td>
  </tr>
</tbody>
</table>

## Function to customize the flowchart

The function `modify_fc` allows the user to customise the created flowchart by modifying its parameters, which are stored in `.$fc`. 

For example, we could fully customise the text in the exclusion box if we wanted to specify the different reasons for exclusion:
 
```{r fig.width = 6, fig.height = 5}
fc <- clinic_patient |> 
  as_fc(label = "Patients included") |> 
  fc_filter(age >= 18 & consent == "Yes", label = "Patients included", show_exc = TRUE) |> 
  fc_modify(~.x |> 
              mutate(
                text = case_when(
                  id == 3 ~ str_glue("Excluded patients:
                                     - {sum(clinic_patient$age < 18)} under-age
                                     - {sum(clinic_patient$consent == 'No')} without a signed consent
                                     "),
                  TRUE ~ text
                )
              )) 

fc |> 
  fc_draw()
```

We could also use `fc_modify()` to change the default _x_ and _y_ coordinates:

```{r fig.width = 6, fig.height = 5}
fc |> 
  fc_modify(~.x |> 
              mutate(
                x = case_when(
                  id == 3 ~ 0.8,
                  TRUE ~ x
                ),
                y = case_when(
                  id == 1 ~ 0.85,
                  id == 2 ~ 0.15,
                  id == 3 ~ 0.5
                )
              )) |> 
  fc_draw()
```

# Export

Once the flowchart has been drawn we can export it to the desired format using `fc_export()`:

```{r eval = FALSE}
clinic_patient |> 
  filter(!is.na(group)) |> 
  as_fc(label = "Patients included") |> 
  fc_split(group) |> 
  fc_draw() |> 
  fc_export("flowchart.png")
```

# Examples

## Example 1

```{r warning = FALSE, fig.width = 7, fig.height = 6}
clinic_patient |> 
  as_fc(label = "Available patients") |> 
  fc_filter(age >= 18 & consent == "Yes", label = "Patients included", show_exc = TRUE) |> 
  fc_split(group) |> 
  fc_filter(n_visits == 2, label = "Two visits available", show_exc = TRUE) |> 
  fc_split(marker_alt, label = c("Marker not altered", "Marker altered")) |> 
  fc_draw()
```

## Example 2

For this example, we will use another built-in dataset called `safo`, which is a randomly generated dataset from the SAFO trial. SAFO is an open-label, multicentre, phase III–IV superiority randomised clinical trial designed to assess whether cloxacillin plus fosfomycin administered during the first 7 days of therapy achieves better treatment outcomes than cloxacillin alone in hospitalised patients with meticillin-sensitive Staphylococcus aureus bacteraemia. In this example, we reproduce the original flowchart of the trial published in Nature Medicine[^1].

[^1]: Grillo, S., Pujol, M., Miró, J.M. et al. Cloxacillin plus fosfomycin versus cloxacillin alone for methicillin-susceptible Staphylococcus aureus bacteremia: a randomized trial. Nat Med 29, 2518–2525 (2023). https://doi.org/10.1038/s41591-023-02569-0

First, we need to do some pre-processing to reproduce the text in the larger boxes:

```{r warning=FALSE, fig.width = 9, fig.height = 7}
# Create labels for exclusion box:
data(safo)

label_exc <- paste(
  c(str_glue("{sum(safo$inclusion_crit == 1 | safo$exclusion_crit == 1 | safo$decline_part == 1, na.rm = T)} excluded:"),
    map_chr(c("inclusion_crit", "decline_part", "exclusion_crit"), ~str_glue("{sum(safo[[.x]] == 1, na.rm = TRUE)} {attr(safo[[.x]], 'label')}")),
    map_chr(4:15, ~str_glue(" -  {sum(safo[[.x]] == 1)} {attr(safo[[.x]], 'label')}"))),
  collapse = "\n")

label_exc <- gsub("exclusion criteria", "exclusion criteria:", label_exc)

safo1 <- safo |> 
  filter(group == "cloxacilin alone", !is.na(reason_pp)) |> 
  mutate(reason_pp = droplevels(reason_pp))

label_exc1 <- paste(
  c(str_glue("{nrow(safo1)} excluded:"),
    map_chr(levels(safo1$reason_pp), ~str_glue(" -  {sum(safo1$reason_pp == .x)} {.x}"))),
  collapse = "\n")

label_exc1 <- str_replace_all(label_exc1, c("resistant" = "resistant\n", "blood" = "blood\n"))

safo2 <- safo |> 
  filter(group == "cloxacilin plus fosfomycin", !is.na(reason_pp)) |> 
  mutate(reason_pp = droplevels(reason_pp))

label_exc2 <- paste(
  c(str_glue("{nrow(safo2)} excluded:"),
    map_chr(levels(safo2$reason_pp), ~str_glue(" -  {sum(safo2$reason_pp == .x)} {.x}"))),
  collapse = "\n")

label_exc2 <- str_replace_all(label_exc2, c("nosocomial" = "nosocomial\n", "treatment" = "treatment\n"))
```

Second, let's create and customise the flowchart using the functions in the package:

```{r warning=FALSE, fig.width = 10, fig.height = 7}
safo |> 
  as_fc(label = "patients assessed for eligibility", text_pattern = "{n} {label}") |> 
  fc_filter(!is.na(group), label = "randomized", text_pattern = "{n} {label}", show_exc = TRUE,
            just_exc = "left", text_pattern_exc = "{label}", label_exc = label_exc, text_fs_exc = 7) |>
  fc_split(group, text_pattern = "{n} {label}") |> 
  fc_filter(itt == 1, label = "included in intention-to-treat\n population", show_exc = TRUE, 
            text_pattern = "{n} {label}", 
            label_exc = "patient did not receive allocated\n treatment (withdrew consent)", 
            text_pattern_exc = "{n} {label}", text_fs_exc = 7) |>
  fc_filter(pp == 1, label = "included in per-protocol\n population", show_exc = TRUE,
            just_exc = "left", text_pattern = "{n} {label}", text_fs_exc = 7) |> 
  fc_modify(
    ~.x |> 
      filter(id != 9) |> 
      mutate(
        text = case_when(id == 11 ~ label_exc1, id == 13 ~ label_exc2, TRUE ~ text),
        x = case_when(id == 3 ~ 0.8, id %in% c(7, 11) ~ 0.15, id == 13 ~ x + 0.035, TRUE ~ x),
        y = case_when(id == 3 ~ y + 0.015, TRUE ~ y)
      )
  ) |> 
  fc_draw()
```

## Example 3

In this example, we will create a flowchart without any dataframe, using the `N =` argument to manually specify the numbers to display in the boxes:

```{r warning=FALSE, fig.width = 7, fig.height = 4.5}
as_fc(N = 300) |> 
  fc_filter(N = 240, label = "Randomized patients", show_exc = TRUE) |> 
  fc_split(N = c(100, 80, 60), label = c("Group A", "Group B", "Group C")) |>
  fc_filter(N = c(80, 75, 50), label = "Finished the study") |> 
  fc_draw()
```

