---
title: "Combine Flowcharts"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 5
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Combine Flowcharts}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(flowchart)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
```

`fc_merge()` and `fc_stack()` allow you to combine different `flowchart`s horizontally or vertically. This is very useful when you need to combine `flowchart`s generated from different `data.frame`s, as shown here. 

# Merge

We can combine different `flowchart`s horizontally using `fc_merge()`. For example, we might want to represent the flow of patients included in the ITT population with the flow of patients included in the PP population.

```{r fig.width = 8}
# Create first flowchart for ITT
fc1 <- safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(itt == "Yes", label = "Intention to treat (ITT)")

fc_draw(fc1)

# Create second flowchart for visits
fc2 <- safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(pp == "Yes", label = "Per protocol (PP)")

fc_draw(fc2)

list(fc1, fc2) |> 
  fc_merge() |> 
  fc_draw()
```

# Stack

We can combine different `flowchart`s vertically using `fc_stack()`. For example, we can combine the same two `flowchart`s vertically instead of horizontally.

```{r warning = FALSE, fig.width = 6, fig.height = 5}
list(fc1, fc2) |> 
  fc_stack() |> 
  fc_draw()
```

We can use the argument `unite = TRUE` to connect two stacked `flowchart`s. For example:

```{r warning=FALSE, fig.width = 6, fig.height = 5}
fc1 <- fc1 |> 
  fc_split(group)

list(fc1, fc2) |> 
  fc_stack(unite = TRUE) |> 
  fc_draw()
```
