---
title: "Flowchart Customization"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 5
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Flowchart Customization}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(flowchart)
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
library(tibble)
```

We can customize the flowchart either with the arguments provided by each function in the process of creating it, or directly in the final output using the `fc_theme()` or `fc_modify()` functions.

# Change function arguments

There are many different arguments in `as_fc()`, `fc_filter()`, `fc_split()`, and `fc_draw()` that allow you to customize the boxes created at each step. See the documentation for these functions for more information. Some examples of how to use these arguments are given at the end of the vignette: [Customization examples].

# Functions to customize the flowchart

## `fc_theme()`

We can use `fc_theme()` to set all parameters for all boxes in a flowchart at once. This is useful if we want to apply the same style to all the boxes. For example, let's customize the following flowchart:

```{r}
safo |> 
  dplyr::filter(!is.na(group)) |> 
  as_fc(label = "Randomized patients") |>
  fc_split(group) |> 
  fc_draw()
```

Let's change the style of all the boxes:

```{r}
safo |> 
  dplyr::filter(!is.na(group)) |> 
  as_fc(label = "Randomized patients") |>
  fc_split(group) |>
  fc_theme(text_pattern_init = "{label}, {n}", text_pattern = "{label}, {n}/{N}", text_fs = 11, text_color = "#324C54", text_fface = 2, bg_fill = "#ADD8E6") |> 
  fc_draw()
```

## `fc_modify()`

The function `fc_modify` allows the user to customise the created flowchart by modifying its parameters, which are stored in `.$fc`. 

For example, let's customize the following flowchart:

```{r fig.width = 6, fig.height = 4}
safo_fc <- safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE) 

safo_fc |> 
  fc_draw()
```

Previous to modifying it, we can use the function fc_view() to inspect the element $fc that we want to change:

```{r}
safo_fc |> 
  fc_view("fc")
```

Let’s customise the text in the exclusion box (id = 3) to specify different reasons for exclusion, and change the x and y coordinate:

```{r}
safo_fc |> 
  fc_modify(
    ~ . |> 
      mutate(
        text = ifelse(id == 3, str_glue("- {sum(safo$inclusion_crit == 'Yes')} not met the inclusion criteria\n- {sum(safo$exclusion_crit == 'Yes')} met the exclusion criteria"), text),
        x = case_when(
          id == 3 ~ 0.75,
          TRUE ~ x
        ),
        y = case_when(
          id == 1 ~ 0.8,
          id == 2 ~ 0.2,
          TRUE ~ y
        )
      )
  ) |> 
  fc_draw()
```

# Customization examples

In these examples, we will explore some of the arguments to customize the following flowchart:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE) |> 
  fc_split(group) |> 
  fc_draw()
```

## Change the flowchart appearance

You can set the background color of each box using `bg_fill=`:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility", width = 0.6, text_fs = 10, text_fface = 2, text_ffamily = "serif", text_padding = 2, bg_fill = "lightgrey") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE, text_color = "white", bg_fill = "darkgreen", text_color_exc = "white", bg_fill_exc = "firebrick") |> 
  fc_split(group, bg_fill = c("darkblue", "purple"), text_color = "white") |> 
  fc_modify(
    ~ . |> 
      mutate(
        y = case_when(
          type == "init" ~ 0.8,
          .default = y
        )
      )
  ) |> 
  fc_draw()
```

You can also alter the background of the canvas behind the flowchart boxes using the `bg_canvas=` argument in `fc_draw`:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility", width = 0.6, text_fs = 10, text_fface = 2, text_ffamily = "serif", text_padding = 2, bg_fill = "lightgrey") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE, text_color = "white", bg_fill = "darkgreen", text_color_exc = "white", bg_fill_exc = "firebrick") |> 
  fc_split(group, bg_fill = c("darkblue", "purple"), text_color = "white") |> 
  fc_modify(
    ~ . |> 
      mutate(
        y = case_when(
          type == "init" ~ 0.8,
          .default = y
        )
      )
  ) |> 
  fc_draw(canvas_bg = "darkgrey")
```

## Add a title to the flowchart

We can add a title to the flowchart using the argument `title=` in the `fc_draw()` function:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE) |> 
  fc_split(group) |> 
  fc_draw(title = "SAFO flowchart")
```

### Add a title to the split

We can also add a title to a split in the flowchart, using the argument `title` in the `fc_split()` function:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE, perc_total = TRUE) |> 
  fc_split(group, perc_total = TRUE, title = "Treatment", bg_fill_title = "skyblue") |> 
  fc_draw()
```

### Percentage with respect to the total rows

We can change the calculation of all percentages in a `flowchart.` By default, percentages are calculated with respect to the box in the previous level. With the argument `perc_total=` we can change it, to calculate it with respect to the initial box with the total number of rows:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE, perc_total = TRUE) |> 
  fc_split(group, perc_total = TRUE) |> 
  fc_draw()
```

### Offset

We can add/remove space to the distance between boxes in a split using the argument `offset`:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE, perc_total = TRUE) |> 
  fc_split(group, offset = 0.1) |> 
  fc_draw()

safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE, perc_total = TRUE) |> 
  fc_split(group, offset = -0.1) |> 
  fc_draw()
```

We can also add/remove space to the distance between the excluded box in a filter using the argument `offset_exc`:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE, offset_exc = 0.1) |> 
  fc_split(group) |> 
  fc_draw()
```

## Change Box Corner Style

We can change the corner style of the flowchart boxes using the `box_corners` argument with `fc_draw`:

```{r fig.width = 6, fig.height = 5}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE) |> 
  fc_split(group) |> 
  fc_draw(box_corners = "sharp")
```


## Use expressions

We can use expressions in the label or the text pattern of each box. Expressions allow you to use bold or italic text without having to change the font of all the box text. 

For example, let's put and expression in the label:

```{r}
safo |> 
  as_fc(label = expression(paste("Patients ", italic("assessed"), " for ", bold("eligibility")))) |>
  fc_draw()
```

Now, let's use it also in the text pattern of the box:

```{r}
safo |>
  as_fc(label = expression(paste("Patients ", italic("assessed"), " for ", bold("eligibility"))), text_pattern = expression(paste("{label}", bold("{n}")))) |>
  fc_filter(filter = chronic_heart_failure == "No",
            label = expression(paste(italic("No chronic"), bold(" heart failure"))),
            label_exc = expression(paste(bold("Chronic"), " heart failure")),
            text_pattern = expression(paste("{label}", N, " = ", bold("{n}"), "/{N}: ", italic("{perc}"), "%")),
            text_pattern_exc = expression(paste("{label}", bold("{n}"), "({perc})%")),
            show_exc = TRUE) |>
  fc_draw()
```

Expressions even allow the use of formulas. For example:

```{r}
as_fc(N = 500, label = expression(paste(y, " = ", alpha, " + ", beta, x)), text_pattern = expression(paste("{label}", bold("{n}")))) |>
  fc_draw()
```

## Split in one group

We can perform an additional split only in one of the groups using the argument `sel_group=`:

```{r fig.width = 6, fig.height = 6}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE) |> 
  fc_split(group) |> 
  fc_split(N = c(50, 60), sel_group = "cloxacillin alone") |> 
  fc_draw()
```

Then, we could also perform a filter in the other group:

```{r fig.width = 6, fig.height = 6}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE) |> 
  fc_split(group) |> 
  fc_split(N = c(50, 60), sel_group = "cloxacillin alone") |> 
  fc_filter(N = 50, sel_group = "cloxacillin plus fosfomycin") |> 
  fc_draw()
```

If we want to select a group in a flowchart with more than two groups we have to supply a vector in `sel_group=` with the desired groups to be selected:

```{r fig.width = 6, fig.height = 7}
safo |> 
  as_fc(label = "Patients assessed for eligibility") |>
  fc_filter(!is.na(group), label = "Randomized", show_exc = TRUE) |> 
  fc_split(group) |> 
  fc_split(N = c(50, 55, 10, 100)) |> 
  fc_filter(N = 60, sel_group = c("cloxacillin alone", "group 2")) |> 
  fc_draw()
```

Previous to modifying it, we can use the function `fc_view()` to inspect the element `$fc` that we want to change:

```{r}
safo_fc |> 
  fc_view("fc")
```

Let's customise the text in the exclusion box (`id = 3`) to specify different reasons for exclusion, and change the _x_ and _y_ coordinate:
 
```{r fig.width = 7, fig.height = 5}
safo_fc |> 
  fc_modify(
    ~ . |> 
      mutate(
        text = ifelse(id == 3, str_glue("- {sum(safo$inclusion_crit == 'Yes')} not met the inclusion criteria\n- {sum(safo$exclusion_crit == 'Yes')} met the exclusion criteria"), text),
        x = case_when(
          id == 3 ~ 0.75,
          TRUE ~ x
        ),
        y = case_when(
          id == 1 ~ 0.8,
          id == 2 ~ 0.2,
          TRUE ~ y
        )
      )
  ) |> 
  fc_draw()
```
